---
title: "Have Middelkoop's predictions come true?"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
knitr::opts_knit$set(root.dir = 'C:/Users/godzi/Documents/eda_rhine')

```


```{r, include=FALSE}
getwd()
```

```{r, include=FALSE}

# Environment setup

library(ggplot2)
library(mapview)
library(data.table)
library(reshape2)
library(sf)
library(moments)

runoff_summary <- readRDS('./data/runoff_summary.rds')
runoff_day <- readRDS('./data/runoff_day.rds')
runoff_month <- readRDS('./data/runoff_month.rds')
runoff_stats <- readRDS('./data/runoff_stats.rds')
runoff_summer <- readRDS('./data/runoff_summer.rds')
runoff_winter <- readRDS('./data/runoff_summer.rds')
runoff_year <- readRDS('./data/runoff_year.rds')

runoff_summary_key <- readRDS('./data/runoff_summary_key.rds')
runoff_day_key <- readRDS('./data/runoff_day_key.rds')
runoff_month_key <- readRDS('./data/runoff_month_key.rds')
runoff_summer_key <- readRDS('./data/runoff_summer_key.rds')
runoff_winter_key <- readRDS('./data/runoff_winter_key.rds')
runoff_year_key <- readRDS('./data/runoff_year_key.rds')

colset_4 <- c("#D35C37", "#BF9A77", "#D6C6B9", "#97B8C2")
n_stations <- nrow(runoff_summary)

```


In 2001 Middelkoop et al. conducted groundbreaking research into the outflow patterns of the Rhine river in Europe. Their models indicated that rising temperatures were leading to a rise in winter discharge due to increased snow-melt, and a decrease in summer discharge due to lower levels of winter snow storage and increase of evapotranspiration in the warmer months.  

Their results were published in the scientific journal *Climatic Change* and influenced policy makers, as well as hydrological projections for the next two decades. 

The full research paper [_Impact of Climate Change on Hydrological Regimes and Water Resources Management in the Rhine Basin_](https://www.researchgate.net/publication/200043966_Impact_of_Climate_Change_on_Hydrological_Regimes_and_Water_Resources_Management_in_the_Rhine_Basin) is available online. 

Now that twenty years have passed, we are going to dive in to the data and examine whether their predictions have come true. 

### Where are we? 

First, let's have a look at the area we will be covering, and at the geographic distribution of the gauging stations from where we are collecting the data. 

```{r, echo = FALSE}

stations <- st_as_sf(runoff_summary, 
                              coords = c("lon", "lat"), 
                              crs = 4326)
stations_map <- mapview(stations, map.types = 'Stamen.TerrainBackground')

stations_map


```
  

From this map we can see that the stations are well-distributed across the river. We can also take this moment to note that the water is flowing from the Alps to the Baltic. This means that the stations in Switzerland are the first and at the very beginning of the catchment area of the Rhine, while the Netherlands is where the it meets the North sea.  

One imporant thing to point out is that along the river there is a lot of variation in terms of topography. The beginning of the catchment is in the Alpine and pre-Alpine areas where the highest and largest differences in altitude are (range of 300m - 1200m). These areas' runoff is most influenced by snow.  
The river then runs through the German Middle Mountain area with altitudes between 150m and 700m, through the scenic Mosel basin, before reaching the lowlands of the Vecht catchment. There are minimal differences in elevation in this stage, and it is where groundwater plays its biggest role for water balance of the catchment.

The stations can therefore be subdivided into three groups - upstream, midstream and downstream. The plot below shows us the average runoff per station, within their flow categories.  


```{r, echo = FALSE}
runoff_month_mean <- runoff_month[, .(value = mean(value)), .(month, sname)]
to_plot <- runoff_month[runoff_summary[, .(sname, category)], on = 'sname']

ggplot(to_plot, aes(x = factor(month), y = value, fill = category, group = month)) +
  geom_boxplot() +
  facet_wrap(~sname, scales = 'free') +
  scale_fill_manual(values = colset_4[c(1, 2, 4)]) +
  xlab(label = "Month") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()


```

We can notice that while the runoff figures for the very first stations in red are in the low hundreds, the very last figures are much higher in the thousands.

This is because the catchment area represents the total area from which precipitation is being collected into the river flow. At the very beginning of the river, at the highest altitudes, the catchment area is small. However by the very last stations, where altitude has decreased, the runoff is at its highest due to the entirety of the river's catchment area being behind the station, collecting discharge to be taken to the North Sea basin. 

To avoid having overwhelmingly large graphs with too much detail, we will be focusing on only three stations moving forward. These are DOMA (Doma/Esr), BASR (Basel, Reinhalle) and KOEL (Koeln), which are each respectively representative of the upstream, mid and downstream categories. 
 
 

### What does the historical data tell us about average seasonal runoff?


We will take a closer look through regression analysis using two different models, loess and linear, to showcase the trends through two different lenses. On the one hand, through the loess model (blue line) we will see in more detail the variations in runoff through time. Our linear model (orange line) creates a general trend line which disregards outliers and highlights the encompassing trends. 


```{r, echo = FALSE}

ggplot(runoff_winter_key, aes(x = year, y = value)) +
  geom_line(col = colset_4[3])+
  geom_point(col = colset_4[3])+
  facet_wrap(~sname, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle('Winter Runoff Trends') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```
  
Middelkoop's research notes that the largest increases in winter flow are to be found in the Alpine area. This is consistent with our data as our graph above shows. Lower temperatures lead to an increase in precipitation in winter months, which in turn lead to a higher rate of runoff. Seeing as the Apine catchment areas are the ones of the highest altitudes, this is where the biggest differences can be seen, in accordance with the paper. 

Further dowmnstream the effects are weakened as these stations are less dependent on snowmelt.  


```{r, echo = FALSE}

ggplot(runoff_summer_key, aes(x = year, y = value)) +
  geom_line(col = colset_4[3])+
  geom_point(col = colset_4[3])+
  facet_wrap(~sname, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle('Summer Runoff Trends') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```
  

Their results also showed that the Alps had the biggest change in summer runoff, with the biggest decrease. This is consistent with our findings above, which show just that in the DOMA catchment. Further downstream the effects are less extreme in the summer. This is due to the decreased snow storage, which has the biggest impact on the Alpine region.   


``` {r, echo = FALSE}
ggplot(runoff_year_key, aes(x = year, y = value)) +
  geom_line(col = colset_4[3])+
  geom_point(col = colset_4[3])+
  facet_wrap(~sname, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle('Yearly Runoff Trends') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```

Our yearly data showcases that the most drastic change in flow is in the Alpine regions, as our seasonal models indicated. While upstream the Rhine is experiencing an overall decrease in flow, the mid and downstream regions are experiencing an increase in average yearly runoff. 

As mentioned previously, the Alpine region is most dependent on snow, so the reduction in snow cover impacts this region the most. Meanwhile the effects of increased precipitation are more obvious for the mid and downstream regions. Additionally the Vecht catchment water balance is somewhat stabilized by groundwater, which isn't as sensitive to immediate changes in temperature and evapotranspiration. 

The graph below showcases precipitation trends for the region:  


```{r, echo = FALSE, message=FALSE}

precip_day <- readRDS('./data/raw/precip_day.rds')
precip_day[, year := year(date)]
precip_day[, month := month(date)]
precip_day <- precip_day[year < 2019]

precip_day[month == 12 | month == 1 | month == 2, season := 'winter']
precip_day[month == 3 | month == 4 | month == 5, season := 'spring']
precip_day[month == 6 | month == 7 | month == 8, season := 'summer']
precip_day[month == 9 | month == 10 | month == 11, season := 'autumn']
precip_day[, season := factor(season, levels = c('winter', 'spring', 'summer', 'autumn'))]

precip_winter <- precip_day[season == 'winter', .(precipitation = sum(value)), by = year]
precip_summer <- precip_day[season == 'summer', .(precipitation = sum(value)), by = year]

year_thres <- 1980
to_plot <- rbind(cbind(precip_winter, season = factor('winter')), 
                 cbind(precip_summer, season = factor('summer'))) 

to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]
to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]

to_plot <- to_plot[year >= 1950]

ggplot(to_plot, aes(year, precipitation, col = season)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = colset_4[c(4, 1)]) +
  geom_smooth(se = F) +
  theme_bw()

```


### Conclusion  

In this summary we simplified the records by focusing on representative stations. Middelkoop notes that there are variations in daily runoff among individual gauging stations, particularly in the Alpine regions where the biggest differences in altitude are. However, even with only selected stations' data we can see that a lot of the indicators match the findings of Middelkoop et al. 

The increase in temperature indicates decreased storage in the form of winter snow and ice, and in turn an increase in winter percipitation, as Middelkoop et al. noted in their conclusions. This in turn leads to an increase in winter runoff, especially in the Alps, but also a decrease in the summer due to fewer depositories. We have already come across the issues that come with these changes in water flow, such as staggeringly low levels in the summers of [2018](https://www.dw.com/en/germany-swelters-under-heat-wave/a-44839919) and [2019](https://rivercruisepassenger.com.au/heatwave-and-low-water-levels-in-europe-may-force-river-cruise-itinerary-changes/), and flooding in the winter of [2020](https://www.general-anzeiger-bonn.de/ga-english/news/rhine-level-in-bonn-is-currently-rising-rapidly_aid-48739343). 

The implications of these changes in water flow are far reaching and severe, with the potential to impact ecology, freshwater availability, commerce, industry and agriculture across several societies. As the articles above show, some of the issues that Middelkoop highlighted in their paper such as inland navigation and flood defense needs have already transpired. In the paper they highlight the importance of approaching these issues with a flexible approach and long-term thinking, as there is still a lot of uncertainty regarding the rate and magnitude of the changes that will be occuring in the future. 

